Wildcard: Spreadsheet-Driven Customization of Web Applications

### 2020-01-10 paper.tex
pdf diagram of airbnb

Spreadsheet-driven customization applies these exact same benefits in a
different context: customizing existing software, rather than building
new software from scratch. We inherit the low floor and high ceiling of
spreadsheets: small tweaks like sorting a list of data in an app can be
done with a single click, while more complex customizations, like
joining in related data from a web API, are supported by a rich formula
system. The tabular data format is versatile, supporting data from a
wide range of applications.

Spreadsheet-driven customization does not require that the application
is actually backed by a spreadsheet---it merely presents a model of the
internal state of the application, which can be exposed by the
application itself or derived from the user interface by third parties.
This means that spreadsheet-driven customization works with existing
applications that people already use. It also allows for exposing a
common data abstraction across applications, enabling users to have a
consistent mechanism for manipulating data from all of their
applications.

@@

To get a sense of how it might feel to use Wildcard, let's see an
example of someone using it to help with booking a trip on the travel
sites Airbnb and Expedia.

The user starts by opening up the Airbnb search listings page to look
for a place to stay. The page looks nice and mostly works well, but is
missing some key features. As mentioned before, this page doesn't allow
the user to sort by price. It also doesn't let them filter by user
rating. Using Wildcard, the user can add these small features, while
leaving the page's design and the rest of its functionality unchanged.

First, the user opens up the Wildcard panel, which shows a table
corresponding to the search results in the page. As they click around in
the table, the corresponding row in the page is highlighted so they can
see the connection between the views.

Then, the user can use standard spreadsheet column header features to
sort the page by price and filter by rating:

Notice how after manipulating the data, the user was able to close the
table view and continue using the website with its original visual
design. The table view offers a way to change the data backing a page,
but does not need to replace the original interface entirely.

Most websites that show lists of data also offer actions that can be
taken on a row in the table, like adding an item to a shopping cart.
Wildcard has the ability to make these actions available in the data
table if the site adapter implements them. The main advantage this
provides is the ability to easily perform an action in bulk across
multiple rows.

For example, it's tedious on Airbnb to click on listings one by one to
add them to a list of favorites. Using Wildcard, we can just select
multiple rows and favorite all of them with one click. Similarly, we can
also open the detailed pages for many listings in new tabs.

Now the user would like to jot down some notes on the pros and cons of
each listing. To do this, they can simply type notes into an additional
column next to each listing, and the notes appear inside the listings in
the original UI. These annotations are also persisted in the browser for
future retrieval.

Wildcard also includes a formula language which enables more
sophisticated tweaks that fetch external data and perform computations.

When traveling without a car, it's nice to evaluate potential places to
stay based on how walkable the surroundings. Using Wildcard formulas, we
can integrate Airbnb with Walkscore, an API that rates the walkability
of any location on a 1-100 scale. When we call the \texttt{WALKSCORE}
formula with the latitude and longitude of the listing, it returns the
walk score as the cell value. Because the cell's contents are injected
into the page, the score also shows up in the page body.

It might seem that Wildcard is only useful on websites that display
lists of tabular data like search results. But in fact, the table
metaphor is flexible enough to represent many types of data. For
example, a form can be represented as a single row, with a column for
each input.

In previous examples the data extracted from the site was marked as
read-only; users cannot change the name or price of an Airbnb listing.
In this next case, the cells are marked as writable, so that changes in
the table are reflected in the form inputs. This becomes useful when
combined with GUI widgets for editing the value of a table cell.

Filling in dates for a flight search typically requires opening up a
separate calendar app to find the right dates, and then manually copying
them into the form. In Wildcard, we can make this easier by providing a
datepicker widget that has privileged access to the user's calendar
information.

@@

\begin{figure}
\centering
\includegraphics{media/architecture-clean.png}
\caption{The architecture of the Wildcard system}
\end{figure}

@@

Scraping data out of the UI has the advantage of working for any data
visible on the page, but it can require a lot of work to convert
unstructured UI to structured data, and sometimes important data is not
shown in the UI. To work around these problems, we have also prototyped
mechanisms for site adapters to observe AJAX requests made by the
browser and extract data directly from JSON responses. (This mechanism
was used in the Airbnb Walkscore example: approximate latitude and
longitude aren't shown in the Airbnb UI, but they are available in AJAX
responses.)

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter: the adapter specifies how to
find the divs representing data rows, and the core platform mutates the
DOM to reflect the table state. The only exception is row actions (like
favoriting an Airbnb listing), which are implemented as regular
Javascript functions running in the context of the page and can mutate
the DOM, simulate clicks on buttons, etc.

@@

In the broadest sense, Wildcard is inspired by systems aiming to make
software into a dynamic medium where end users frequently create and
modify software to meet their own needs, rather than only consuming
applications built by programmers. These systems include Smalltalk
\autocite{kay1977}, Hypercard \autocite{hypercard2019}, Boxer
\autocite{disessa1986}, Webstrates \autocite{klokmose2015}, and
Dynamicland \autocite{victor}. (The project's name Wildcard comes from
the internal pre-release name for Hypercard, which doubly inspired our
work by promoting both software modification by end users and the ideas
behind the Web.)

While similar in high-level goals, Wildcard employs a different solution
strategy from these projects: whereas they generally require building
new software from scratch for that environment, Wildcard instead aims to
maximize the malleability of already existing software. This approach
has the pragmatic benefit of immediately being useful in a much broader
context, although it also requires working within rigid constraints. We
also see Wildcard growing beyond its current bounds and becoming more
similar in the future to these other systems. With substantial future
work, Wildcard could grow from merely being a platform for tweaking
existing software into a platform for building new software from
scratch, designed from the ground up for end user tweakability.

More narrowly, Wildcard builds on three areas of existing work: web
augmentation, spreadsheet-based app development, and web scraping. Our
contribution in this work is to combine these areas in a new way:
extracting structured data from a website and directly exposing it to
the user as a live data table, with the end goal of modifying an
existing application. We think this combination builds on valuable ideas
and learnings from each of the areas, while also overcoming some of the
downsides of each.

@@

``Sloppy programming'' \autocite{little2010} tools like Chickenfoot
\autocite{bolin2005} and Coscripter \autocite{leshed2008} enable users
to create scripts that perform actions like filling in text boxes and
clicking buttons, without directly interacting with the DOM. Users
express the desired page elements in natural, informal terms
(e.g.~writing ``the username box'' to represent the textbook closest to
the label ``username''), and then using heuristics to determine which
elements most likely match the user's intent. This approach allows for
expressing a wide variety of commands with minimal training, but it also
has downsides. It is difficult to know whether a command will
consistently work over time (in addition to changes to the website,
changes to the heuristics can also cause problems), and it is not easy
for users to discover the space of possible commands.

@@

\emph{Copy of the sloppy contrast that used to be earlier:}

Layering a structured view on top of an existing application is not the
only option available. ``Sloppy programming'' systems like Chickenfoot
\autocite{bolin2005} and CoScripter \autocite{leshed2008} forego the
intermediate layer of structure, instead allowing users to create
scripts in an informal language and then perform fuzzy pattern matching
to find elements in the DOM. For example, to find a button after a
textbox in Chickenfoot, the user could type
\texttt{click(find(“button\ just\ after\ textbox”))}. These designs
allow for expressing a wide range of operations, but they don't
explicitly indicate what operations are possible---the user can only see
the original page and imagine the possibilities. In contrast, Wildcard
provides affordances that clearly suggest the availability of certain
actions (e.g.~sorting, editing a cell, adding a column with a derived
value), especially to users who are familiar with spreadsheets. In
addition to giving users more certainty about whether a modification is
possible, these affordances might improve discoverability and give users
new ideas for things to try. ``Sloppy programming'' does give end-users
greater access to modify the apps that they use, but does not seem to
provide them with a deeper understanding of those apps or how they
relate to each other, so the apps are more likely to remain siloed.

@@

Our prototype of Wildcard is still in early development; there are still
many limitations to resolve and open questions to explore.

The most important question is whether the combination of features
provided by Wildcard can be combined in enough useful ways to make the
system worth using in practice. While initial demos are promising, we
need to develop more site adapters and use cases to more fully assess
this question. We plan to continue privately testing the system with our
own needs, and to eventually deploy the tool publicly, once the API is
stable enough and can support a critical mass of sites and use cases. We
also plan to run usability studies to evaluate and improve the design of
the tool.

Here are a few of the largest limitations in the current system:

\begin{itemize}
\tightlist
\item
  Wildcard only extracts data visible on the page, which means that
  subsequent pages of results in paginated lists are not included in the
  table. (Sifter \autocite{huynh2006} uses techniques that might help
  get around this.)
\item
  There is no mechanism for end users to express imperative workflows
  (e.g.~``click this button, and then\ldots{}''); they can only write
  formulas that return data and then inject the resulting data into the
  page. While this makes the system simpler, it also may exclude many
  valuable use cases. We may add a system for expressing workflows like
  this, although it's not obvious how it would fit together with the
  existing table view. Gneiss \autocite{chang2014} contains some
  mechanisms for writing spreadsheet formulas which can handle behaviors
  like reloading data in response to a user pressing a button, which
  might prove helpful.
\item
  Wildcard's data model only shows a single table at a time, without any
  notion of relationships between tables. A richer data model with
  foreign keys might be necessary to support certain use cases. For
  designing a tabular interface on top of a richer data model, we could
  learn from the interface of Airtable which shows related objects in a
  table cell, or add nested rows as used in other systems
  \autocite{mccutchen2016,bakke2016}.
\item
  Only allowing site adapters to be manually coded means that the number
  of supported sites is limited. We plan to explore abstractions that
  make it as easy as possible for programmers to efficiently create new
  robust adapters, as well as integrating automated heuristics into the
  adapter creation process.
\end{itemize}

@@

We live in a time of pessimism about the effects of the internet and the
degree of centralized control wielded by large corporations. We think
that one promising approach to this problem is to enable people to have
deeper control over the Web software they use every day. The Web offers
the potential of an open foundation, but even the success of browser
extensions has only empowered users within the limited confines of the
ideas that programmers have decided to build.

In this paper, we have presented Wildcard, a browser-embedded
programming system that maps websites to a structured data table,
enabling end users to modify their behavior. We hope that it contributes
to making the web into a more dynamic medium that users can mold to
their own needs.

We plan to continue developing the system and to eventually deploy it as
an open-source tool. To receive future updates on Wildcard and
notifications about a public release, \href{}{sign up for the email
newsletter}.

We are also looking for private beta testers. If you have an idea for
how you might want to use Wildcard, \href{mailto:glitt@mit.edu}{please
contact us}. We would love to collaborate on building site adapters and
formulas to support your use case.

>>>>

Spreadsheet-driven customization applies these exact same benefits in a
different context: customizing existing software, rather than building
new software from scratch. We inherit the low floor and high ceiling of
spreadsheets: small tweaks like sorting a list of data in an app can be
done with a single click, while more complex customizations, like
joining in related data from a web API, are supported by a rich formula
system. The tabular data format is versatile, supporting data from a
wide range of applications. Spreadsheet-driven customization does not
require that the application is actually backed by a spreadsheet---it
merely presents a model of the internal state of the application, which
can be exposed by the application itself or derived from the user
interface by third parties. This means that spreadsheet-driven
customization works with existing applications that people already use.
It also allows for exposing a common data abstraction across
applications, enabling users to have a consistent mechanism for
manipulating data from all of their applications.

@@

To get a sense of how it might feel to use Wildcard, let's see an
example of someone using it to help with booking a trip on the travel
sites Airbnb and Expedia. These demos are best viewed in the online
version of this paper, which includes videos:
\url{https://geoffreylitt.com/wildcard}

\begin{figure}
\hypertarget{fig:airbnb-demo}{%
\centering
\includegraphics{media/airbnb-demo-300dpi.png}
\caption{Using Wildcard to augment the Airbnb page for searching travel
accommodations}\label{fig:airbnb-demo}
}
\end{figure}

The user starts by opening up the Airbnb search listings page to look
for a place to stay. The page looks nice and mostly works well, but is
missing some key features. As mentioned before, this page doesn't allow
the user to sort by price. It also doesn't let them filter by user
rating. Using Wildcard, the user can add these small features, while
leaving the page's design and the rest of its functionality unchanged
(shown in fig.~\ref{fig:sorting}).

After manipulating the data, the user can close the table view and
continue using the website with its original visual design. The table
view offers a way to change the data backing a page, but does not need
to replace the original interface entirely.

Most websites that show lists of data also offer actions that can be
taken on a row in the table, like adding an item to a shopping cart.
Wildcard has the ability to make these ``row actions'' available in the
data table if the site adapter implements them. The main advantage this
provides is the ability to easily perform a row action in bulk across
multiple rows.

For example, it's tedious on Airbnb to click on listings one by one to
add them to a list of favorites. Using Wildcard, the user can just
select multiple rows and favorite all of them with one click. Similarly,
we can also open the detailed pages for many listings in new tabs.
(Shown in fig.~\ref{fig:actions})

\begin{figure}
\hypertarget{fig:expedia-demo}{%
\centering
\includegraphics{media/expedia-demo-300dpi.png}
\caption{Using Wildcard to augment the Expedia page for booking a
flight}\label{fig:expedia-demo}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig:architecture}{%
\centering
\includegraphics{media/architecture-clean.png}
\caption{The architecture of the Wildcard
system}\label{fig:architecture}
}
\end{figure}

@@

Sometimes important data is not shown in the UI, making it impossible to
scrape from the DOM. To address this, we have also prototyped mechanisms
for site adapters to observe AJAX requests made by the browser and
extract data directly from JSON responses. This mechanism was used to
implement the Airbnb Walkscore example, since latitude and longitude
aren't shown in the Airbnb UI, but they are available in AJAX responses.
We also plan to add the ability for site adapters to scrape data across
multiple pages in paginated lists of results (as explored in prior work
\autocite{huynh2006}).

The site adapter also needs to support the reverse direction: sending
updates from the table to the original page. Most DOM manipulation is
not performed directly by the site adapter; instead, the adapter
specifies how to find the divs representing data rows, and the core
platform mutates the DOM to reflect the table state. The only exception
is row actions (like favoriting an Airbnb listing), which are
implemented as imperative Javascript functions that can can mutate the
DOM, simulate clicks on buttons, etc.

@@

\hypertarget{malleable-software}{%
\subsection{Malleable software}\label{malleable-software}}

In the broadest sense, Wildcard is inspired by systems aiming to make
software into a dynamic medium where end users frequently create and
modify software to meet their own needs, rather than only consuming
applications built by programmers. These systems include Smalltalk
\autocite{kay1977}, Hypercard \autocite{hypercard2019} , Boxer
\autocite{disessa1986}, Webstrates \autocite{klokmose2015}, and
Dynamicland \autocite{victor}.\footnote{The project's name Wildcard
  comes from the internal pre-release name for Hypercard, which doubly
  inspired our work by promoting both software modification by end users
  and the ideas behind the Web.}

While similar in high-level goals, Wildcard employs a different solution
strategy from these projects: whereas they generally require building
new software from scratch for that environment, Wildcard instead aims to
maximize the malleability of already existing software. This approach
has the pragmatic benefit of being immediately useful in more
situations, although it also requires working within more rigid
constraints.

With substantial future work, Wildcard could become more similar to
these other projects, growing from a platform for tweaking existing
software into a platform for building new software from scratch. This
would likely end up resembling existing tools for building
spreadsheet-driven applications (discussed more below), but with an
extra focus on customizability by end users of the software.

@@

``Sloppy programming'' \autocite{little2010} tools like Chickenfoot
\autocite{bolin2005} and Coscripter \autocite{leshed2008} enable users
to create scripts that perform actions like filling in text boxes and
clicking buttons, without directly interacting with the DOM. Users
express the desired page elements in natural, informal terms
(e.g.~writing ``the username box'' to represent the textbook closest to
the label ``username''), and then using heuristics to determine which
elements most likely match the user's intent. This approach allows for
expressing a wide variety of commands with minimal training, but it also
has downsides. It is difficult to know whether a command will
consistently work over time (in addition to changes to the website,
changes to the heuristics can also cause problems), and it is not easy
for users to discover the space of possible commands
\autocite{little2010}.

@@

@@

There are still many open questions about spreadsheet-driven
customization, which we hope to answer through targeted development and
usage of the Wildcard prototype.

The most important question is whether the computational model can
support a large variety of useful customizations. While initial demos
are promising, we need to develop more site adapters and use cases to
explore this further. We plan to continue privately testing the system
with our own needs, and to eventually deploy the tool publicly, once the
API is stable enough and can support a critical mass of sites and use
cases. We also plan to run usability studies to evaluate and improve the
design of the tool.

We suspect that two areas of the current model may prove too limiting.
First, Wildcard's data model only shows a single table at a time,
without any notion of relationships between tables. A richer data model
with foreign keys might be necessary to support certain use cases. For
designing a tabular interface on top of a richer data model, we could
learn from the interface of Airtable which shows related objects in a
table cell, or add nested rows as used in other systems
\autocite{mccutchen2016,bakke2016}.

Second, there is no mechanism for end users to express imperative
workflows (e.g.~``click this button, and then\ldots{}''); they can only
write formulas that return data and then inject the resulting data into
the page. If it proves useful, we might consider adding a system for
expressing workflows like this, although it would substantially alter
the computational model and it's not clear how it would fit together
with the existing table view.

Another open question is how efficiently site adapters can be created
for real websites. By making adapters for sites with different types of
data in different domains, we plan to explore how automated heuristics
can speed up the adapter creation process, and to create abstractions
that make it easier for programmers to efficiently create new robust
adapters.

@@

In this paper, we have presented \emph{spreadsheet-driven
customization}, a way to make software customization more accessible to
end users by showing the data inside the application in a spreadsheet.
We hope that this technique contributes to making the Web into a more
dynamic medium that users can mold to their own needs.

We plan to continue developing the Wildcard prototype and to eventually
deploy it as an open-source tool. To receive future updates on Wildcard
and notifications about a public release, \href{}{sign up for the email
newsletter}.

We are also looking for private beta testers. If you have an idea for
how you might want to use Wildcard, please
\href{mailto:glitt@mit.edu}{get in touch}. We would love to hear about
your needs and help find ways to use Wildcard to solve them.
